---
title: "GBS - Second Test"
author: "Mathias Lherbier"
date: "2024-01-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(terra)
library(tidyverse)
```

## Introduction

The goal of this analysis is to review a piece of code that computes the area of the land use class 'Natural Forest' in km². 

Input data are two rasters of similar dimensions, resolution and extent :
- The FLII_earth_resampled raster contains information on inferred or observed forest cover. 
- The Major_land_cover_palm_agroforestry_corrected raster provides land cover data, split in 4 classes (cropland, forestry, pasture and urban area) and natural area cover.


## Data pre-processing
Data is loaded using the rast function.
```{r}

temp_year_file_folder <- './'
Major_land_cover_palm_agroforestry_corrected <- terra::rast('./Major_land_cover_palm_agroforestry_corrected.tif')
FLII_earth_resampled <- terra::rast('./flii_earth_resampled.tif')

```

We will first be cropping rasters in order to reduce computation time as testing goes on.
A sixth of the entire map will be used for our tests meaning that computation times would approximately be six times longer with the entire dataset.
```{r}

crop_extent <-c(-30,30,-15,15)
FLII_earth_resampled_c <- crop(FLII_earth_resampled, crop_extent)
Major_land_cover_palm_agroforestry_corrected_c <- crop(Major_land_cover_palm_agroforestry_corrected,crop_extent)

```

We will be evaluating computation time using the function system.time() from the base package.

# Base function
We first evaluate the function provided in the test:
```{r}
GBS_dev_test2 <- function(temp_year_file_folder,
                          Major_land_cover_palm_agroforestry_corrected,
                          FLII_earth_resampled) {


  #Natural forest on cells where FLII > 9600 ####
  #(values have been multiplied by 1000 according to read me file of FLII data: in the paper > 9.6)

  forestry_natural_rate_filename <- glue::glue("{temp_year_file_folder}/forestry_natural_rate.tif")
  
  if (file.exists(forestry_natural_rate_filename) == FALSE) {
    
    forestry_natural_rate_intermediate <- terra::lapp(c(Major_land_cover_palm_agroforestry_corrected,
                                                        FLII_earth_resampled),
                                                      fun = function(x, y) ifelse((x == 3000 & y >= 9600), 1, 0),
                                                      filename = glue::glue("{temp_year_file_folder}/forestry_natural_rate_intermediate.tif"),
                                                      overwrite = TRUE)
    
    #Reclassify NA values into 0 (otherwise loss of 9Mkm² of forests worldwide)
    forestry_natural_rate <- terra::lapp(c(forestry_natural_rate_intermediate),
                                         fun = function(x) ifelse(x %in% c(NA, NaN, "NA", "NaN"),
                                                                  0,
                                                                  x),
                                         filename = forestry_natural_rate_filename)
    
    
  } else {
    forestry_natural_rate <- rast(forestry_natural_rate_filename)
  }
  
  
  #Natural forest areas (km²) per GLOBIO country ####
  Major_land_cover_area_filename <- glue::glue("{temp_year_file_folder}/terrestrial_LU_major_land_cover_area.tif")
  
  if (file.exists(Major_land_cover_area_filename) == FALSE) {
    Major_land_cover_area <- cellSize(Major_land_cover_palm_agroforestry_corrected,
                                      unit = "km",
                                      filename = Major_land_cover_area_filename)
  } else {
    Major_land_cover_area <- rast(Major_land_cover_area_filename)
  }
  
  #Multiply rate x area ####
  forestry_natural_rate_x_area_filename <- glue::glue("{temp_year_file_folder}/forestry_natural_rate_x_area.tif")
  
  if (file.exists(forestry_natural_rate_x_area_filename) == FALSE) {
    forestry_natural_rate_x_area <- terra::lapp(c(forestry_natural_rate,
                                                  Major_land_cover_area),
                                                fun = function(x, y) x * y,
                                                filename = forestry_natural_rate_x_area_filename,
                                                overwrite = TRUE)
  } else {
    forestry_natural_rate_x_area <- rast(forestry_natural_rate_x_area_filename)
  }
  
}

 #Testing computation time
 system.time(GBS_dev_test2(temp_year_file_folder,
                           Major_land_cover_palm_agroforestry_corrected_c,
                           FLII_earth_resampled_c)) # 144s in my test run
 
 #Saving output
 forestry_natural_rate_x_area <- rast('./forestry_natural_rate_x_area.tif')
 
```


# Introducing dplyr package and the vectorized function is.na()
In this second version of the function, some minor code changes have been applied to improve readability : declaring filenames first, removing negation in the condition.
In terms of performance, we now test using the case_when() from the dplyr package instead of ifelse().
Since case_when() returns NAs when the condition is not met, we can also use the is.na() function in the second treatment.
```{r}
GBS_dev_test2_case <- function(temp_year_file_folder,
                          Major_land_cover_palm_agroforestry_corrected,
                          FLII_earth_resampled) {
  
    forestry_natural_rate_filename <- glue::glue("{temp_year_file_folder}/forestry_natural_rate.tif")
    forestry_natural_rate_intermediate_filename <-glue::glue("{temp_year_file_folder}/forestry_natural_rate_intermediate.tif")
    Major_land_cover_area_filename <- glue::glue("{temp_year_file_folder}/terrestrial_LU_major_land_cover_area.tif")
    forestry_natural_rate_x_area_filename <- glue::glue("{temp_year_file_folder}/forestry_natural_rate_x_area.tif")
  
  #Natural forest on cells where FLII > 9600 ####
  #(values have been multiplied by 1000 according to read me file of FLII data: in the paper > 9.6)
  if (file.exists(forestry_natural_rate_filename)) {
        forestry_natural_rate <- rast(forestry_natural_rate_filename)
  } else {
    forestry_natural_rate_intermediate <- terra::lapp(c(Major_land_cover_palm_agroforestry_corrected,
                                                        FLII_earth_resampled),
                                                      fun = function(x, y) case_when(
                                                        x == 3000 & y >= 9600 ~ 1),
                                                      filename = forestry_natural_rate_intermediate_filename,
                                                      overwrite = TRUE)
    
    #Reclassify NA values into 0 (otherwise loss of 9Mkm² of forests worldwide)
    forestry_natural_rate <- terra::lapp(c(forestry_natural_rate_intermediate),
                                         fun = function(x) ifelse(is.na(x),0,x),
                                         filename = forestry_natural_rate_filename)
    }
  
  
  #Natural forest areas (km²) per GLOBIO country ####
  if (file.exists(Major_land_cover_area_filename)) {
    Major_land_cover_area <- rast(Major_land_cover_area_filename)
  } else {
    Major_land_cover_area <- cellSize(Major_land_cover_palm_agroforestry_corrected,
                                      unit = "km",
                                      filename = Major_land_cover_area_filename)
    }
  
  #Multiply rate x area ####
  if (file.exists(forestry_natural_rate_x_area_filename)) {
    forestry_natural_rate_x_area <- rast(forestry_natural_rate_x_area_filename)
  } else {
    forestry_natural_rate_x_area <- terra::lapp(c(forestry_natural_rate,
                                                  Major_land_cover_area),
                                                fun = function(x, y) x * y,
                                                filename = forestry_natural_rate_x_area_filename,
                                                overwrite = TRUE)
    }
  
}

 #Testing computation time
 system.time(GBS_dev_test2_case(temp_year_file_folder,
                           Major_land_cover_palm_agroforestry_corrected_c,
                           FLII_earth_resampled_c)) # 89s in my test run
 
 #Saving output
 forestry_natural_rate_x_area_case <- rast('./forestry_natural_rate_x_area.tif')

```


# Taking advantage of the Raster Dataset format from terra package
Instead of using the function c() to merge the rasters, we will now try using the function sds() that creates a SpatRasterDataset from the two rasters.
```{r}
GBS_dev_test2_case_sds <- function(temp_year_file_folder,
                               Major_land_cover_palm_agroforestry_corrected,
                               FLII_earth_resampled) {
  
    forestry_natural_rate_filename <- glue::glue("{temp_year_file_folder}/forestry_natural_rate.tif")
    forestry_natural_rate_intermediate_filename <-glue::glue("{temp_year_file_folder}/forestry_natural_rate_intermediate.tif")
    Major_land_cover_area_filename <- glue::glue("{temp_year_file_folder}/terrestrial_LU_major_land_cover_area.tif")
    forestry_natural_rate_x_area_filename <- glue::glue("{temp_year_file_folder}/forestry_natural_rate_x_area.tif")

  
  #Natural forest on cells where FLII > 9600 ####
  #(values have been multiplied by 1000 according to read me file of FLII data: in the paper > 9.6)
  if (file.exists(forestry_natural_rate_filename)) {
    forestry_natural_rate <- rast(forestry_natural_rate_filename)
  } else {
    forestry_natural_rate_intermediate <- terra::lapp(sds(list(Major_land_cover_palm_agroforestry_corrected,
                                                        FLII_earth_resampled)),
                                                      fun = function(x, y) case_when(
                                                        x == 3000 & y >= 9600 ~ 1),
                                                      filename = forestry_natural_rate_intermediate_filename,
                                                      overwrite = TRUE)
    
    #Reclassify NA values into 0 (otherwise loss of 9Mkm² of forests worldwide)
    forestry_natural_rate <- terra::lapp(c(forestry_natural_rate_intermediate),
                                         fun = function(x) ifelse(is.na(x),0,x),
                                         filename = forestry_natural_rate_filename)
    }
  
  #Natural forest areas (km²) per GLOBIO country ####
  if (file.exists(Major_land_cover_area_filename)) {
    Major_land_cover_area <- rast(Major_land_cover_area_filename)
  } else {
    Major_land_cover_area <- cellSize(Major_land_cover_palm_agroforestry_corrected,
                                      unit = "km",
                                      filename = Major_land_cover_area_filename)
    }
  
  #Multiply rate x area ####
  if (file.exists(forestry_natural_rate_x_area_filename)) {
    forestry_natural_rate_x_area <- rast(forestry_natural_rate_x_area_filename)
  } else {
    forestry_natural_rate_x_area <- terra::lapp(sds(list(forestry_natural_rate,
                                                  Major_land_cover_area)),
                                                fun = function(x, y) x * y,
                                                filename = forestry_natural_rate_x_area_filename,
                                                overwrite = TRUE)
    }
}

 #Testing computation time
 system.time(GBS_dev_test2_case_sds(temp_year_file_folder,
                                    Major_land_cover_palm_agroforestry_corrected_c,
                                    FLII_earth_resampled_c)) # 62s in my test run
 
 #Saving output
 forestry_natural_rate_x_area_case_sds <- rast('./forestry_natural_rate_x_area.tif')

```

We now examine if all outputs are equal : 
```{r}
all.equal(forestry_natural_rate_x_area,forestry_natural_rate_x_area_case) #TRUE

all.equal(forestry_natural_rate_x_area,forestry_natural_rate_x_area_case_sds) # Only the component name is different (lyr1 vs lyr.1) but data is iso between the two outputs.

```

All results being equal, we would favor the third iteration of the function for improved computation times.

# Converting results to a data frame

The function as.data.frame() from the terra package can be used to turn the results into a data frame.
The option xy being TRUE, the coordinates are included in the data frame.

```{r}
results <- as.data.frame(forestry_natural_rate_x_area_case_sds, xy=T)

```

If need be, results could be split into chunks, then converted to data frames and merged into one data frame to optimize performance.


#Potential improvements

To potentially decrease computation time, we could consider parallelizing certain operations.
The function terra::lapp has an argument called cores built in that allows for parallel processing but it did not prove to be efficient in my tests. 
A specific package called parallel also allows to process in parallel, while declaring the number of cores to be used in each cluster.


Similarly to what I have done for my performance tests, we could process data in chunks to reduce computation times.
Data could be split in chunks using the crop() function and eventually re-assembled using cbind() or rbind() if we decided to work on strips.


Whatever the strategy we will end up using, it will be necessary to test its performance on a smaller dataset to make sure it fits our needs and does not alter the results.